---
title: "S610 - Final Project"
author: "Gokcen Buyukbas, Jared Roush"
date: "12/9/2020"
output: html_document
---
#Latex needed
 Let $X=[x_{ij}]$ denote the $n \times d$ configuration matrix associated with $x_1,\dots{},x_n\in \mathbb{R}^d$. Let\\

$$d_{ij}=d_{ij}(X)=\|x_i-x_j\|=\left[ \sum_{l=1}^{d}(x_{il}-x_{jl})^2 \right]^{1/2}$$
denote the Euclidean distance between $x_i$ and $x_j$. Gien an nxn dissimilarity matrix $\Delta=[\delta_{ij}]$, the weighted raw stress criterion is\\
$$\sigma(X)=\sum_{i\leftarrow \rightarrow j}w_{ij} \left[ d_{ij}(X)-\delta_{ij}  \right]^2=\frac{1}{2}\sum_{i,j=1}^nw_{ij} \left[ d_{ij}(X)-\delta_{ij}  \right]^2$$

 where $w_{ij}=w_{ji}\geq 0$ is the weight assigned to approximating $\delta_{ij}=\delta_{ji}$ with $d_{ij}=d_{ji}$



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(usedist) #dist_make(data matrix, fcn)
library(plot3D)
library(plot3Drgl)
library(scatterplot3d) 
library(rgl) #scatter3d()
source("Useful_functions.R")
```

$\sigma(x) = $


## Isomap function
## Takes a data set, a localization parameter eps, desired dimension d, 
##    returns: coordinates of a representation  in d-dim, 


```{r}
isomap = function(X, eps=NULL, k=NULL, d){
  n = nrow(X)
  Delta_E = as.matrix(dist(X, diag = TRUE, upper = TRUE)) #Find pairwise distances 
  if (eps != 'NULL'){
    Delta_E[Delta_E  > eps] = Inf # Replace values greater than eps with infinity
  }
  Delta_G = Delta_E
  ## Find the shortest path between the points i&j in the epsilon graph
  E = matrix(0,nrow=n,ncol=n) 
  m = 1
  while (m < n-1){
    for (i in 1:n){
      for (j in 1:n){
        E[i,j] = min(Delta_G[i,]+Delta_G[,j])
      }
    }
    Delta_G = E
    m = 2*m
  }
  R0 = cmdscale(Delta_G, d, eig=TRUE)$points
  for(i in 1:50){
    R0 = mds.guttman.eq(R0, Delta_G)
  }
  R1 = R0
  return(R1)
}
```



### Examples ###

# Swiss Roll
```{r}
n0 = 300

swissroll = data.frame(x = numeric(n0),
                       y = numeric(n0),
                       z = numeric(n0))
for (i in 1:n0){
  u = runif(1, 0,1)
  v = runif(1, 0,1)
  swissroll$x[i] = 3^u*cos(8*u)
  swissroll$y[i] = 3^u*sin(8*u)
  swissroll$z[i] = v
}
```
```{r}

plot(swissroll)

scatterplot3d(swissroll$z, swissroll$x, swissroll$y)
scatter3d(swissroll$x, swissroll$y, swissroll$z, asp=1)

swiss2 = isomap(swissroll, eps=0.7, d=2)
plot(swiss2)

```

# Rectangular annulus

```{r}
n1 = 100
R <- matrix(0,nrow=1,ncol=2)
for (i in 1:n1) {
  x1 <- runif(1,min=0,max=3)
  y1 <- runif(1,min=0,max=1)
  x2 <- runif(1,min=0,max=1)
  y2 <- runif(1,min=1,max=3)
  x3 <- runif(1,min=2,max=3)
  y3 <- runif(1,min=1,max=3)
  x4 <- runif(1,min=0,max=3)
  y4 <- runif(1,min=3,max=4)
  R <- rbind(R,c(x1,y1),c(x2,y2),c(x3,y3),c(x4,y4))
}
plot(R, main="Original data")

R1 = isomap(R, eps=0.5, d=2)
plot(R1)
```

# Spiral

```{r}
n2 = 200
S = data.frame(x = numeric(n2),
               y = numeric(n2),
               z = numeric(n2))
for (i in 1:n2){
  t = runif(1, 0, 1)
  S$x[i] = cos(8*t)
  S$y[i] = sin(8*t)
  S$z[i] = t
}

plot(S)
scatterplot3d(S)

S2 = isomap(S, eps=0.5, d=2)
plot(S2)
```

# On Torus
```{r}
R=1
r=0.5
n3 = 200
Torus = data.frame(x = numeric(n3),
                   y = numeric(n3),
                   z = numeric(n3))
for (i in 1:n3){
  u = runif(1, 0, pi)
  v = runif(1, 0, pi)
  Torus$x[i] = (R+r*cos(u))*cos(v)
  Torus$y[i] = (R+r*cos(u))*sin(v)
  Torus$z[i] = r*sin(u)
}

scatterplot3d(Torus)
plot(Torus)

T2 = isomap(Torus, eps=0.5, 2)
plot(T2)
```
